Material Methods

### 2020-02-05 doc/MatMeth.Rmd
working on the chrm class proportions sections, other small edits
start working on MatMethods again
reviewing MatMethods doc adding notes based on new analysis

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The Mus musculus PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were cryo-derived from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp).
The related species of *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jax labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken.

 All mice were housed UW-Madison Biotech and MSC facilities - following the protocols.  A breeding colony of wild derived 
*Mus musculus* mice sampled from Gough Island (GI)is maintained at UW Veterinary school facilities.
 Mice were feed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison.

Over the course of data collection some breeding colonies  mice were moved from facilities. Additionally the GI strain was kept at a separate facility.  (evidence for environmental (room) effect on (MLH1 / SC lengths / a variety of meiotic traits was tested for. We observed no effect, Supplemental figure))

@@

The vast majority of oocyte data was collected from neonate litters/mice between 5 to 48 hours. This approach maintain breeding pairing and still result in prophase oocytes (cite timeline of oocytes in neonates). (differences were assessed between embryo samples and neonates?). Precise staging of embryos by checking copulary plugs was difficult/not feasible in many of these wild derived strains due to their behavior. Embryonic samples were collected when pregnancy noted in females, embryos were staged based on (X table markers).

@@

The same staining protocol was applied to spermatocyte and oocyte spreads.
The staining protocol was based on X with some adaptations, and same as previously described in Peterson et al 2019.

Explain list recipes for DMC1 since these haven't been listed before

@@

For MLH1 pachytene characterization, Cells with a full karyotype (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes), distinct foci, and intact bivalents were included for quantification.
Reprocessing in Photoshop (cite). Image file names were anonymized before manual scoring of MLH1 or DMC1 foci.

@@

@@

@@

We chose to use a **mixed model framework** to interpret the patterns of variation, built using the lmer4 package (cite) in R (v3.5.2)[@Rstudio].

@@

@@

>>>>

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The Mus musculus strains PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were generated from frozen embryos purchased  from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp) and rederived by the UW-Biotech centre (Madison UW).
The related murid species *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jackson labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken.

 All mice were housed UW-Madison Biotech and MSC facilities - following the protocols.  A breeding colony of wild derived *Mus musculus* mice sampled from Gough Island (GI)is maintained at UW Veterinary school facilities.
 Mice were fed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison.

Over the course of data collection some breeding colonies  mice were moved from facilities. Additionally the GI strain was kept at a separate facility.  (evidence for environmental (room) effect on (MLH1 / SC lengths / a variety of meiotic traits was tested for. Effects on average MLH1 counts were compared -- for strains which were kept at mulitple institutions. ((We observed no effect, Supplemental figure))

@@

The vast majority of oocyte data was collected from neonate litters/mice between 5 to 48 hours old. This approach maintain breeding pairing and still result in prophase oocytes (cite timeline of oocytes in neonates). (differences were assessed between embryo samples and neonates?).

Precise staging of embryos via copulary plugs was not feasible in many of these wild derived strains due to their behavior. Embryonic samples were collected when pregnancy noted in females, embryos were staged based on (X table markers).

@@

SYCP3 was used to mark  visualize the Axial element of the synaptonemal complex. Sicne all mouse chromosomes are telocentric/acrocentric, CREST was used to indicate the centromere -- indicate the side of the bivalent with the centromere.

the genome wide recombination rate was charaecterized / estimated with staining of MLH1 antibody, 
The same staining protocol was applied to spermatocyte and oocyte spreads.
The staining protocol was based on X with some adaptations, and same as previously described in [@peterson_2019].

Precursurs to COs, DSBs were quantified with the DMC1 staning antibody. Due to resource constraints -- only a subset of strains were stained for this marker. Only spermatocyte spreads produced enough good cells for DMC1 quantification.

@@

For MLH1 pachytene characterization, cells with a full karyotype (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes), distinct foci, and intact bivalents were included for quantification.
Image file names were anonymized before manual scoring of MLH1 or DMC1 foci.

@@

Foci counts/numbers were manually scored.

Single bivalent features were measured both manual and with an automated image analysis software [@Peterson_2019]

DMC1 -- leptotene and sygotene cells were identified by SC morphology (by degree of synapsis) (cite another paper?). foci counted manually, error ccalulated from a subset of cells counted twice (supplement)

total number of DMC1 foci per cell were reported.

@@

DMC1 -- Ttest

@@

The mean, variance and coefficient of variance for MLH1 counts per cell were calculated for each mouse (in R). these mouse level summary statistics were analyzed -- within our model frameworks.

We chose to use a **mixed model framework** to interpret the patterns of variation, built using the lmer4 package (cite) in R (v3.5.2)[@Rstudio]. (lme)

@@

### Single bivalet quant

Single bivalent chromosomes were isolated from cytology cell images as discribed in (X ). the single bivalents were 'curated' (removed errors in the segmentation algorithm)

to account for effects of CO number and physical chromosome size -- the data was also analyzed after spliting up by chromosome class and in long and short chromosome bins.

- Split up 'classes' of bivalents depending on the number of foci -- to tests/ comparisons more comprable.

- The performance of the segmentaion algorithm was assessed by counting the number of objects segmented per single cell image (expected 19 or 20 in males and females respectively). All automated bivalent data was manually curated before analysis.

- whole cell bivalent measures (measures of all bivalents ), were compiled by mixed automated bivalent measures with hand measurements. (Since the error in automated measures was so low)

**SC length**

The axial element (AE) of the SC -- for single bivalents (synapsesd homolog pairs), was quantified as the skeletonized (single pixel width line) --

-comparison to human measures previously in permyscus species -- and in musculuc (shown in the supplement)

- The long bivalent and short bivalent data sets were compiled by calculating the 1st and 4th quartille of SC lengths for each cell with at least 16 bivalent measures. Bivalents from each cell were assigned into the long and short bins based on these SC Length cut offs.

**CO interference**

- how we choose to measure interference, the average distance between2 foci on same bivalent.

- downside to using the gamma distribution -- not the primary meteric for CO interference.

- Also pooling intervals for categories and fitting a gamma distribution (in supplement)

**rec landscape** **1CO normalized position**

- how why measure the recombination landscape -- coohse to charaecterize the rec landscaoe bby isloating the 1CO bivalents; since there are known sexual dimorphic features of this and since bivalents with more than 1 foci will be primarily influenced by CO interference. We look at both the raw and

the distance to the telomere and centromere end of the SC shape were measured -- for all / pooled bivalents

@@

-Mixed models and glm models were built

-tests for fixed effects
-tests of random effects in mixed models was LTR of exactRLRT function in X package.

single bivalent analysis

# Numbers of mice cells ect for each method?

above belongs more in results
